Copulas and their potential for ecology

### 2019-10-18 Paper.Rmd
small edits on final read through, part done

Plots using normalized ranks were proposed in the Introduction as being conceptually
similar to copulas. We here explain the connection. As stated above, the copula associated
with a random variable $(X,Y)$ is the joint distribution function of the random variable
$(F_X(X),F_Y(Y))$. Given a sample $(x_i,y_i)$, $i=1,\ldots,n$, from $(X,Y)$, let $\hat{F}_X$ and
$\hat{F}_Y$ be the empirical cdfs associated with the $x_i$ and $y_i$, respectively.
These are step functions of $x$ and $y$, respectively, that start at $0$ for low values of 
$x$ and $y$ and jump by $1/n$
at each of the data points; for large $n$ they approximate the cdfs $F_X$ and $F_Y$. 
But $\hat{F}_X(x_i)$ equals the normalized rank of $x_i$ times
$n/(n+1)$, and likewise $\hat{F}_Y(y_i)$ equals the normalized rank of $y_i$ times
$n/(n+1)$. So the normalized rank in turn approximates the empirical cdf and therefore 
the cdf. Thus the normalized rank pairs $(u_i,v_i)$ for $i=1,\ldots,n$ can be 
regarded as approximate samples from the random variable $(F_X(X),F_Y(Y))$, 
which is the random variable associated with the copula of $(X,Y)$.
The scatterplot of $v_i$ versus $u_i$ can be used to infer aspects of copula 
structure (see section on methods for Q1 below).

@@

We addressed Q3 by exploring, using both data and models, three hypothesized
ecological consequences of non-normal copula structure and asymmetric tail
associations. As for Q2, models used are intended as initial explorations, only,
of whether copula structure may possibly have a meaningful influence on important ecological
phenomena. Therefore only simple models are used, and comprehensive explorations of the 
sensitivity of results to model strucuture and parameters are left for future work.

The hypothesis was presented in the Introduction that
the distribution (through time) of a spatially 
averaged quantity should be influenced by 
dependencies between the local quantities being averaged, including their
copula structure and tail associations. 
To refine this hypothesis, suppose an ecological variable 
$E_i(t)$ is measured at locations $i=1,\ldots,N$ and times $t=1,\ldots,T$, and
the spatial mean $\sum_i E_i(t)/N$ is of
interest. The $E_i(t)$ could be, for instance, local abundances of a pest or 
exploited species, or local fluxes of a greenhouse gas. If  $E_i$ and $E_j$ are 
associated primarily in their right tails for most location pairs $i$ and $j$, then
exceptionally large values tend to occur at the same time in most locations.
We hypothesize that this 
will tend to increase the skewness of the distribution 
of the spatial mean. Similarly, left-tail associations between local variables
should tend to decrease skewness. Strong positive skewness of the 
spatial-mean time series corresponds to "spikiness" of that time series,
which corresponds to instability through time. 
The spatial-mean time series and its skewness may be quantities of 
principal importance for 
pest or resource abundance, for which extreme values (spikes) 
in the spatial mean may have large effects.

@@

To explore the influence of copula structure and tail associations on spatial Taylor's
law, we carried out a series of simulations that generated sets of population 
matrices that had 
distinct copula structure between locations but that 
were otherwise statistically similar. For $k=1,\ldots,1000$
and X representing a Clayton, normal, Frank, or a survival Clayton copula family,
we generated $N \times n$ population matrices $m^{(\text{X},k)}$ with the following properties:
1) for any given values of $j$ and $k$, the values 
$\{ m_{t,j}^{(\text{X},k)} : t=1,\ldots,N  \}$ were exactly the same, as unordered
sets, for all X, i.e., the same actual population values were used for location $j$ 
in simulation $k$, regardless of X;
2) for any given $k$ and $j_1$ and $j_2$ such that $j_1 \neq j_2$, the Spearman correlations (computed through time)
$\cor_t(m_{t,j_1}^{(\text{X},k)},m_{t,j_2}^{(\text{X},k)})$ were the same, to within sampling
variation, i.e., correlations through time of populations in two locations were the same,
up to sampling variation, regardless of X;
3) the copula for the dependence between $m_{t,j_1}^{(\text{X},k)}$ and $m_{t,j_2}^{(\text{X},k)}$ was X.
We used $N=50$ and $n=25$. Details of how these matrices were generated are in 
Appendices \ref{SM-Consequence_TL_simulation} and \ref{SM-sect:copsurrognd}.
Thus for each $k$, the poulation matrices $m^{(\text{X},k)}$
for X taking the values Clayton, normal, Frank, and survival Clayton
were statistically similar except for the copula structure between locations, which was
X. Thus comparing how Taylor's law may manifest differently for $m^{(\text{X},k)}$
for different values of X constitutes a test of the influence of copula
structure on Taylor's law. We selected X equal to the Clayton, normal, Frank, and survival 
Clayton copulas to explore a range of tail association patterns.

@@

\noindent Results were consistent with the hypothesis that the skewness
of a spatial-average time series is influenced by tail associations
between the local quantities being averaged. 
For datasets that
exhibited highly asymmetric tail associations in earlier analyses (green spruce 
aphid abundance data had stronger lower-tail association, and, 
respectively,  leaf-curling plum 
aphid first flight data had stronger upper-tail association, Table 
\ref{tab_multivar_summary}), skewness of the spatial average 
was less than (respectively, greater than) a significant fraction of 
surrogate skewness values 
(Fig. \ref{fig_skewness_spearman}A,B). For datasets with moderately stronger 
lower than upper-tail 
dependence (\emph{Ceratium furca} abundance and methane data), 
skewness of the spatial
average showed a non-significant or marginally significant tendency 
toward being less than surrogate skewnesses (Fig. \ref{fig_skewness_spearman}C,D). 
Thus copula 
structure and asymmetric tail associations are important for spatially averaged 
quantities. This result is represented in Fig. \ref{MasterFigure}
as the solid box around "Instability/skewness of mean or total 
time series" and the solid arrows labeled "X".

@@

Also consistent with hypothesis, copula structure had a substantial effect on
Taylor's law. 
Taylor's law was strongly influenced, and was often even invalidated in its assumptions of 
linearity and homoskedasticity, by non-normal copula structure. For normal copula structure 
(i.e., for simulations that gave normal-copula dependence between populations
in different locations), $p$-values for the linearty
and homoskedasticity tests were roughly uniformly distributed across replicate 
simulations and Taylor's law appeared visually to be a reasonable approximation
of the $\log(v_g)$ versus $\log(m_g)$ relationship (Fig. \ref{fig_spTL}B,E,F). 
Furthermore, quadratic coefficients
and curvature values were close to $0$ (Fig. \ref{fig_spTL}J,K), and root mean squared
errors from the linear regression were relatively small (Fig. \ref{fig_spTL}G).
But linearity or homoskedasticity were violated more frequently for non-normal 
copula structure
(Fig. \ref{fig_spTL}A,C,D,E,F); quadratic coefficients and curvatures were frequently 
non-zero (Fig. \ref{fig_spTL}J,K); and root mean squared
errors from the linear regression were much higher (Fig. \ref{fig_spTL}G). Slopes 
and intercepts of the linear regression were also strongly affected by copula
structure (Fig. \ref{fig_spTL}H,I), though some of the effect here was because
linear regressions do not always adequately represent the $\log(v_g)$ versus $\log(m_g)$
relationship when copula structure was not normal.
Thus our results substantiated the hypothesis, at least for the models we used,
that Taylor's law can be influenced by copula structure and asymmetric
tail associations. This is represented in Fig. \ref{MasterFigure} by the 
arrow labelled "Z".

@@

<!--what we did and why is it important to study?-->
We showed that non-normal copula structure and asymmetric tail associations are 
common across multiple sub-disciplines
in ecology, although these facets of ecological data are only occasionally 
taken into account [@Valpine2014; @Anderson2018; @Popovic2018]. We hypothesized 
mechanisms that may cause non-normal copula structures and asymmetric tail associations;
we discuss below how commonly some of our
mechanisms may operate. 
We also hypothesized and demonstrated important consequences of 
non-normal copula structure and asymmetric tail associations for the field of ecology. 
For instance, the skewness
of a spatial-average time series is influenced by asymmetric tail associations between its 
constituent time series: predominantly right-tail-associated local time series 
can lead to "spiky" spatially averaged time series, with large outbreaks.
Thus tail associations could have implications for exploited and pest species.
Extinction risk and Taylor's law can also be altered by tail association patterns across space
in environmental variables. 
In our view, our results make it reasonable to suggest 
that a more comprehensive understanding 
of many ecological phenomena may be possible if a
complete copula characterization of dependence is employed. 
Copula approaches are
well developed [@nelsen2006_copula; @joe2014_dependence; @MaiScherer2017], statistically, 
and have been introduced in accessible formats [@Anderson2018; @Genest2007]. Furthermore, 
open-source computer implementations exist (e.g., the `copula` and 
`VineCopula` packages in the R programming language). Ecologists can apply these tools
immediately. We created several interrelated randomization procedures
(Appendices \ref{SM-surrog_test}, \ref{SM-Surrog_multi}, \ref{SM-sect:copsurrognd}) 
that build upon existing copula methods.

<!--tropical ecosystems-->
The approaches we have demonstrated should apply equally well to 
data from tropical or temperate ecosystems.
There is no reason to expect that non-normal copula structure and 
asymmetric tail dependence should be special properties of datasets from temperate 
regions. The mechanisms we have proposed of non-normal copula structure and asymmetric tail 
dependence seem equally likely to apply anywhere. For instance, the Moran effect, which underpins 
one of our proposed mechanisms (Fig. \ref{MasterFigure}B), is a standard mechanism
that occurs whenever environmental variables influence populations. And Liebig's
law and nonlinear environmental influences on ecosystems, which underpin 
another one of our proposed mechanisms (Fig. \ref{MasterFigureA}), are widely
demonstrated paradigms that apply to all biomes.

@@

<!--para on how likely one is to get copula structure via the Moran mechanism-->
There are also reasons to hypothesize that 
our Moran mechanism (Fig. \ref{MasterFigure}, B) 
may operate commonly: Moran effects are common [@Liebhold2004], and 
non-normal copula structures and asymmetric tail associations are often found in 
environmental variables. If intense meteorological events are 
also widespread, then
environmental variables associated with these events should take extreme
values simultaneously across large spatial areas, producing tail dependence in measurements
made through time at different locations. Non-extreme values may instead
be associated with local phenomena, and therefore may be
less correlated across large areas.
@Serinaldi2008 examined the spatial dependence 
of rainfall in Central Italy; Gumbel or Student 2-copulas 
were candidates for modelling dependence, and 
neither of these is a normal copula.
A long-term study (1950–2014) in the Loess Plateau of China [@She2018]
showed that a Gumbel copula effectively modeled 
the spatial dependence of drought variables; the Gumbel copula demonstrates 
asymmetric tail associations.
Bivariate copula analysis was also used in forecasting the 
co-occurrence of extreme events (flood or drought) 
over the North Sikkim Himalayas using spatial datasets [@Goswami2018].

>>>>

Plots using normalized ranks were proposed in the Introduction as being conceptually
similar to copulas. We here explain the connection. As stated above, the copula associated
with a random variable $(X,Y)$ is the joint distribution function of the random variable
$(F_X(X),F_Y(Y))$. Given a sample $(x_i,y_i)$, $i=1,\ldots,n$, from $(X,Y)$, let $\hat{F}_X$ and
$\hat{F}_Y$ be the empirical cdfs associated with the $x_i$ and $y_i$, respectively.
These are step functions of $x$ and $y$, respectively, that start at $0$ for low values of 
$x$ and $y$ and jump by $1/n$
at each of the data points. Therefore $\hat{F}_X(x_j)$ is the rank of $x_j$ in
the set $\{x_i : i=1,\ldots,n\}$, dividided by $n$, 
and analogously $\hat{F}_Y(y_j)$ is the rank of $y_j$ in the set 
$\{y_i : i=1,\dots,n\}$, divided
by $n$. For large $n$, $\hat{F}_X(x_i)$ and $\hat{F}_Y(y_i)$ 
approximate the cdfs $F_X$ and $F_Y$, respectively. 
But $\hat{F}_X(x_i)$ equals the normalized rank of $x_i$ times
$n/(n+1)$, and likewise $\hat{F}_Y(y_i)$ equals the normalized rank of $y_i$ times
$n/(n+1)$. So the normalized rank in turn approximates the empirical cdf and therefore 
the cdf. Thus the normalized rank pairs $(u_i,v_i)$ for $i=1,\ldots,n$ can be 
regarded as approximate samples from the random variable $(F_X(X),F_Y(Y))$, 
which is the random variable associated with the copula of $(X,Y)$.
The scatterplot of $v_i$ versus $u_i$ can be used to infer aspects of copula 
structure (see section on methods for Q1 below).

@@

We addressed Q3 by exploring, using both data and models, three hypothesized
ecological consequences of non-normal copula structure and asymmetric tail
associations. As for Q2, the models we used are intended as initial explorations, only,
of whether copula structure may have a meaningful influence on important ecological
phenomena. Therefore only simple models are used, and comprehensive explorations of the 
sensitivity of results to model strucuture and parameters are left for future work.

The hypothesis was presented in the Introduction that
the distribution (through time) of a spatially 
averaged quantity should be influenced by 
dependencies between the local quantities being averaged, including their
copula structure and tail associations. 
To make this hypothesis more precise, suppose an ecological variable 
$E_i(t)$ is measured at locations $i=1,\ldots,N$ and times $t=1,\ldots,T$, and
the spatial mean $\sum_i E_i(t)/N$ is of
interest. The $E_i(t)$ could be, for instance, local abundances of a pest or 
exploited species, or local fluxes of a greenhouse gas. If  $E_i$ and $E_j$ are 
associated primarily in their right tails for most location pairs $i$ and $j$, then
exceptionally large values tend to occur at the same time in most locations.
We hypothesize that this 
will tend to increase the skewness of the distribution 
of the spatial mean. Similarly, left-tail associations between local variables
should tend to decrease skewness. Strong positive skewness of the 
spatial-mean time series corresponds to "spikiness" of that time series,
i.e., occassional very large values, 
which corresponds to instability through time. 
The spatial-mean time series and its skewness may be quantities of 
principal importance for 
pest or resource abundance, for which extreme values (spikes) 
in the spatial mean may have large effects.

@@

To explore the influence of copula structure and tail associations on spatial Taylor's
law, we carried out a series of simulations that generated sets of population 
matrices that had 
distinct copula structure between locations but that 
were otherwise statistically similar. For $k=1,\ldots,1000$
and X representing a Clayton, normal, Frank, or a survival Clayton copula family,
we generated $N \times n$ population matrices $m^{(\text{X},k)}$ with the following properties:
1) for any given values of $j$ and $k$, the values 
$\{ m_{t,j}^{(\text{X},k)} : t=1,\ldots,N  \}$ were exactly the same, as unordered
sets, for all X, i.e., the same actual population values were used for location $j$ 
in simulation $k$, regardless of X;
2) for any given $k$ and $j_1$ and $j_2$ such that $j_1 \neq j_2$, the Spearman correlations (computed through time)
$\cor_t(m_{t,j_1}^{(\text{X},k)},m_{t,j_2}^{(\text{X},k)})$ were the same, to within sampling
variation, i.e., correlations through time of populations in two locations were the same,
up to sampling variation, regardless of X;
3) the copula for the dependence between $m_{t,j_1}^{(\text{X},k)}$ and $m_{t,j_2}^{(\text{X},k)}$ was  from the family X.
We used $N=50$ and $n=25$. Details of how these matrices were generated are in 
Appendices \ref{SM-Consequence_TL_simulation} and \ref{SM-sect:copsurrognd}.
Thus for each $k$, the poulation matrices $m^{(\text{X},k)}$
for X taking the values Clayton, normal, Frank, and survival Clayton
were statistically similar except for the copula structure between locations, which was
X. Thus comparing how the $log(v_g)$ versus $\log(m_g)$ relationship may
manifest differently for $m^{(\text{X},k)}$
for different values of X constitutes a test of the influence of copula
structure on Taylor's law. We selected X equal to the Clayton, normal, Frank, and survival 
Clayton copulas to explore a range of tail association patterns.

@@

\noindent Empirical results were consistent with the hypothesis that the skewness
of a spatial-average time series is influenced by tail associations
between the local quantities being averaged. 
For datasets that
exhibited highly asymmetric tail associations in earlier analyses (green spruce 
aphid abundance data had stronger lower-tail association, and, 
respectively,  leaf-curling plum 
aphid first flight data had stronger upper-tail association, Table 
\ref{tab_multivar_summary}), skewness of the spatial average 
was less than (respectively, greater than) a significant fraction of 
surrogate skewness values 
(Fig. \ref{fig_skewness_spearman}A,B). For datasets with moderately stronger 
lower than upper-tail 
dependence (\emph{Ceratium furca} abundance and methane data), 
skewness of the spatial
average showed a non-significant or marginally significant tendency 
toward being less than surrogate skewnesses (Fig. \ref{fig_skewness_spearman}C,D). 
Thus copula 
structure and asymmetric tail associations are important for spatially averaged 
quantities. This result is represented in Fig. \ref{MasterFigure}
as the solid box around "Instability/skewness of mean or total 
time series" and the solid arrows labeled "X".

@@

Also consistent with hypothesis, copula structure had a substantial effect on
Taylor's law for the models we considered. 
Taylor's law was strongly influenced, and was often even invalidated in its assumptions of 
linearity and homoskedasticity, by non-normal copula structure. For normal copula structure 
(i.e., for simulations that gave normal-copula dependence between populations
in different locations), $p$-values for the linearty
and homoskedasticity tests were roughly uniformly distributed across replicate 
simulations and Taylor's law appeared visually to be a reasonable approximation
of the $\log(v_g)$ versus $\log(m_g)$ relationship (Fig. \ref{fig_spTL}B,E,F). 
Furthermore, quadratic coefficients
and curvature values were close to $0$ (Fig. \ref{fig_spTL}J,K), and root mean squared
errors from the linear regression were relatively small (Fig. \ref{fig_spTL}G).
But linearity or homoskedasticity were violated more frequently for non-normal 
copula structure
(Fig. \ref{fig_spTL}A,C,D,E,F); quadratic coefficients and curvatures were frequently 
non-zero (Fig. \ref{fig_spTL}J,K); and root mean squared
errors from the linear regression were much higher (Fig. \ref{fig_spTL}G). Slopes 
and intercepts of the linear regression were also strongly affected by copula
structure (Fig. \ref{fig_spTL}H,I), though some of the effect here was because
linear regressions do not always adequately represent the $\log(v_g)$ versus $\log(m_g)$
relationship when copula structure was not normal.
Thus our results substantiated the hypothesis, at least for the models we used,
that Taylor's law can be influenced by copula structure and asymmetric
tail associations. This is represented in Fig. \ref{MasterFigure} by the 
arrow labelled "Z".

@@

<!--what we did and why is it important to study?-->
We showed that non-normal copula structure and asymmetric tail associations are 
common across multiple sub-disciplines
in ecology, although these facets of ecological data are only occasionally 
taken into account [@Valpine2014; @Anderson2018; @Popovic2018]. We hypothesized 
mechanisms that may cause non-normal copula structures and asymmetric tail associations;
we discuss below how commonly some of our
mechanisms may operate. 
We also hypothesized and demonstrated important consequences of 
non-normal copula structure and asymmetric tail associations for the field of ecology. 
For instance, the skewness
of a spatial-average time series is influenced by asymmetric tail associations between its 
constituent time series: predominantly right-tail-associated local time series 
can lead to "spiky" spatially averaged time series, with large outbreaks; and 
predominantly left-tail-associated local time series can lead to spatially
averaged time series showing accentuated "crashes".
Thus tail associations could have implications for pest and exploited species.
Extinction risk and Taylor's law can also be altered by tail association 
patterns across space. 
In our view, our results make it reasonable to suggest 
that a more comprehensive understanding 
of many ecological phenomena may be possible if a
complete copula characterization of associations between variables is employed. 
Copula approaches are
well developed, statistically [@nelsen2006_copula; @joe2014_dependence; @MaiScherer2017], 
and have been introduced in accessible formats [@Anderson2018; @Genest2007]. Furthermore, 
open-source computer implementations exist (e.g., the `copula` and 
`VineCopula` packages in the R programming language). Ecologists can apply these tools
immediately. We created several interrelated randomization procedures
(Appendices \ref{SM-surrog_test}, \ref{SM-Surrog_multi}, \ref{SM-sect:copsurrognd}) 
that build upon existing copula methods.

<!--tropical ecosystems-->
The approaches we have demonstrated should apply equally well to 
data from tropical or temperate ecosystems.
There is no reason to expect that non-normal copula structure and 
asymmetric tail associations should be special properties of datasets from temperate 
regions. The mechanisms we have proposed of non-normal copula structure and asymmetric tail 
associations seem equally likely to apply anywhere. 
For instance, the Moran effect, which underpins 
one of our proposed mechanisms (Fig. \ref{MasterFigure}, B), is a standard mechanism
that occurs whenever environmental variables influence populations. And Liebig's
law and nonlinear environmental influences on ecosystems, which underpin 
another one of our proposed mechanisms (Fig. \ref{MasterFigure}, A), are widely
demonstrated phenomena that apply in all biomes.

@@

<!--para on how likely one is to get copula structure via the Moran mechanism-->
There are also reasons to hypothesize that 
our Moran mechanism (Fig. \ref{MasterFigure}, B) 
may operate commonly: Moran effects are common [@Liebhold2004; @Sheppard2016; 
@Defriez2017_ocean; @Defriez2017_land], and 
non-normal copula structures and asymmetric tail associations are often found in 
environmental variables. If intense meteorological events are 
also widespread, then
environmental variables associated with these events should take extreme
values simultaneously across large spatial areas, producing tail associations in measurements
made through time at different locations. Non-extreme values may instead
be associated with local phenomena, and therefore may be
less correlated across large areas.
@Serinaldi2008 examined the spatial dependence 
of rainfall in Central Italy. Gumbel or Student 2-copulas 
were candidates for modelling dependence, and 
neither of these is a normal copula.
A long-term study (1950–2014) in the Loess Plateau of China [@She2018]
showed that a Gumbel copula effectively modeled 
the spatial dependence of drought variables. The Gumbel copula demonstrates 
asymmetric tail associations.
Bivariate copula analysis was also used in forecasting the 
co-occurrence of extreme events (flood or drought) 
over the North Sikkim Himalayas using spatial datasets [@Goswami2018].

<!--***DAN: made it to here 2019 10 18-->
